\documentclass[a4paper, 11pt, english]{article}
\usepackage[top=3.17cm, bottom=3.27cm, left=3.5cm, right=3.5cm]{geometry}
%\setlength{\parindent}{2.5em}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
{\renewcommand{\arraystretch}{1.3}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{latexsym}
\usepackage{amsmath,amsthm,amssymb,amsfonts}
\usepackage{eucal}
\usepackage[english]{babel}
\usepackage{tikz}
\usepackage{epsfig}
\usepackage{amssymb, graphicx, amsmath, amsthm}
\usepackage{calrsfs}
\usepackage{todonotes}
\usepackage[hidelinks]{hyperref}
\usepackage{bm}
\usepackage{appendix}
\usepackage{multicol}
\usepackage[useregional]{datetime2}
\usepackage{listings}
\usepackage{tikz-cd}

\RequirePackage{graphicx}
\RequirePackage{hyperref}
\newcommand{\parm}{\mathord{\color{black!33}\bullet}}

\newcommand{\breath}{\vspace{6pt plus 2pt minus 1pt}\noindent}

\newcommand{\pattern}[4]{
  \raisebox{0.6ex}{
  \begin{tikzpicture}[scale=0.3, baseline=(current bounding box.center), #1]
    \foreach \x/\y in {#4}
      \fill[NE-lines] (\x,\y) rectangle +(1,1);
    \draw (0.01,0.01) grid (#2+0.99,#2+0.99);
    \foreach \x/\y in {#3}
      \filldraw (\x,\y) circle (5pt);
  \end{tikzpicture}}
}

\newcommand{\patternrule}{ \mapsto \!}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{openproblem}[theorem]{Open Problem}
\newtheorem*{openproblem*}{Open Problem}
\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem*{remark*}{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem*{example*}{Example}

\newcommand{\eps}{\epsilon}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\etal}{et~al.}
\newcommand{\Sym}{S}
\newcommand{\Pow}{\mathcal{P}}
\newcommand{\Ocal}{\mathcal{O}}
\newcommand{\emptyword}{\varepsilon}
\newcommand{\id}{\mathrm{id}}
\newcommand{\rid}{\overline{\id}}
\newcommand{\from}{\leftarrow}
\newcommand{\eq}{\leftrightarrow}
\newcommand{\eqstar}{\stackrel{*}{\leftrightarrow}}
\newcommand{\tostar}{\stackrel{*}{\to}}
\newcommand{\toplus}{\stackrel{+}{\to}}
\newcommand{\fromstar}{\stackrel{*}{\from}}
\newcommand{\nf}[1]{{#1\!\downarrow}}
\newcommand{\floor}[1]{\lfloor#1\rfloor}
\newcommand{\bracket}[1]{[\mskip1mu #1\mskip1mu]}
\DeclareMathOperator{\olap}{\mathcal{O}}
\DeclareMathOperator{\dom}{dom}
\DeclareMathOperator{\img}{img}
\DeclareMathOperator{\des}{des}
\DeclareMathOperator{\maj}{maj}
\DeclareMathOperator{\ovl}{ovl}

\begin{document}

\title{Enumerating Hertzsprung equivalence classes defined by partitions of the symmetric group of degree 3}
\author{Bjarki Hartjenstein Gunnarsson}
\date{\today}
\maketitle

\begin{abstract}
    Hertzsprung patterns are permutation patterns in which both positions and values are required to
    be adjacent. A family of sets of Hertzsprung patterns induce equivalence classes by
    a pattern-replacement equivalence. In this paper we introduce a method that allows us 
    to enumerate these equivalence classes for some families. We manage to find the joint
    distrubution of equivalence classes defined by a majority of the set partitions of $\Sym_3$.

    \breath \emph{Keywords}: permutation pattern, pattern-rewriting system
\end{abstract}

\section{Introduction}
Permutation patterns are sub-permutations which we can match with parts of other
permutations, under various constrains. 
Classical permutation patterns do not have any constraints on their matches but they can
be constrained such that the positions or values of the matches must be adjacent.
Hertzsprung patterns constrain both positions and values to be adjacent. 

%%% skip mentioning mesh patterns altogether?
These constraints can be formalized by mesh patterns \cite{claesson:2011}.

In this demonstration we make pattern matches bold.  Let us treat $132$ as a classical
pattern, then we would have the following matches in the permutation $15432$.
\[
    \bm{154}32
\]
\[
    \bm{1}54\bm{32}
\]
If we look again at the same pattern and permutation, but assume that the
pattern match is constrained so that positions must be adjacent then we have the match
\[
    \bm{154}32.
\]
Again, looking at at the same pattern and permutation, but assuming that the
pattern is constrained so that the values must be adjacent we have the following
match:
\[
    \bm{1}54\bm{32}.
\]

Hertzsprung patterns constrain the match so both values and positions must be
adjacent. The example permutation above avoids the Hertzsprung pattern $132$, that is, it does not have a match
within the permutation. But as a showcase, given the pattern $132$ and the permutation $1243$ we
have the Hertzsprung match
\[
    1\bm{243}.
\]

From this point onward, whenever we mention patterns we are referring to
Hertzsprung patterns.

Finally, a Hertzsprung pattern that matches the beginning of a permutation is
called the (Hertzsprung) \emph{prefix} of that permutations. In a similar way, a pattern
that matches the end of a permutation we call a (Hertzsprung) \emph{suffix}.

\subsection{Equivalences, equivalence classes and motivation}
We can define an equivalence relation using patterns. For a permutation with a certain pattern
match, we can look at a transposition of that pattern match into another pattern will result in an
equivalent permutation. 

Say we consider $123$ and $321$ being equivalent. Then given the
permutation with a pattern match $123$ marked in bold:
\[
  5\bm{234}1
\]
we rearrange the match into the equivalent pattern $321$ and get
\[
  5\bm{432}1.
\]
As a consequence of the patterns $123$ and $321$ being equivalent the permutations $52341$ and
$54321$ land in the same equivalence class. That is, the two permutations are equivalent up to the
\emph{pattern-replacement equivalence} induced by the equivalence of $123$ and $321$.

We put patterns that can replace each other into sets, and then collect those sets together. 
\[
    \{ \{\alpha, \beta, \gamma, \dots \}, \{\lambda, \mu, \nu, \dots \}, \dots
    \}.
\]
and we have what we call in this paper a family of equivalences.

The collection of sets above is a shorthand for
\[
    \{ \{ \alpha, \beta \}, \{ \alpha, \gamma \}, \dots, \{ \lambda, \mu \}, \{
        \lambda, \nu \}, \dots \}
\] 
which we call the canonical form. The shorthand notation is equivalent to the canonical form since
we are working with equivalence relations and they are transitive.
That is, we can replace the pattern $\beta$ with $\alpha$ and then $\alpha$ with $\gamma$, for
example, to get that $\beta$ and $\gamma$ are in the same equivalence class.

Note that there are sometimes more than two different canonical forms for a family of sets. For each 
set of equivalent patterns we choose one common member and then link every other element to the set to that common member
with unordered pairs, in the example above we choose $\alpha$ and $\lambda$ as the common members
but they could be any other member of the respective sets.

Again, a collection of sets of equivalent patterns induces a pattern-replacement equivalence.
To put it in more formal language, a family of equivalent-pattern sets $J$ will induce an equivalence relation
which we call $\equiv_J$. We say $u \pi v \equiv_J u \rho v$ if $\{ \pi, \rho \} \in J$, with $u,v
\in \Sym$.

For example, if we say $J = \{ \{ 132, 321 \} \}$ then we consider the patterns $132$ and $321$ to
be equivalent by the relation $\equiv_J$. That will e.g. result in
\[
    1\bm{243}5 \equiv_J 1\bm{432}5
\]
or
\[
  \bm{132}4\bm{765} \equiv_J \bm{321}4\bm{576}
\]

An equivalence relation on $\Sym = \cup_{n \geq 1} \Sym_n$ splits each $\Sym_n$
into a set of equivalence classes which we want to enumerate. 

Linton et al. had already made some headway on enumerating equivalence classes of patterns that constrain
positions to be adjacent on one hand, and patterns that constrain values be adjacent on the other.
Not much work had been done on enumerating equivalence classes of patterns that constrain both
positions and values, Hertzsprung patterns, until the spring of 2021 when Claesson published a
preprint of his article \cite{claesson:2021} that gives a semi-automatic way of enumerating equivalence classes of the
last type.

To enumerate these different equivalence classes we will introduce Claesson's 
pattern-rewriting systems. They are the only part of the method given by Claesson that requires
some manual work. The work involves finding systems that fulfil a certain property, namely
\emph{termination}.

\subsection{Pattern-rewriting systems}
A \emph{pattern-rewriting system} is a subset $R \subset \cup_{n} (\Sym_n \times \Sym_n)$.
We use the notation $\pi \patternrule \rho$ for $(\pi, \rho)$ when referring
to elements of a pattern-rewriting system $R$. 

Juxtaposition of permutations signifies a direct sum. If $\pi=123$ and
$\tau=321$ then $\pi\tau=123654$. The left argument is left unchanged while the
right one is pushed up by the length of the left argument.

As for variable naming, we use the variables $a$ and $b$ for whole permutations, 
the variables $\pi$, $\rho$, $\tau$ and $\sigma$ for patterns.
Finally we use the variables $u$ and $v$ for sub-permutations, permutations that make up larger
permutations.

We can make a pattern-rewriting system induce a \emph{rewrite relation} on $\Sym$ in the following way: We say
$a \to_R b$ for $a, b \in \Sym_k$ if there exists $\pi \patternrule \rho \in R$ so
we can write $a = u \pi v$ and $b = u \rho v$, where
$u$ and $v$ are in $\Sym$.  Note that the relation is
length-preserving. For example, if $R = \{123 \patternrule 321\}$, then $1234 \to_R 3214$. 

We denote the domain of $R$ by $\dom(R) = \{\pi : \pi \patternrule
\rho \in R\}$. For the image of $R$ we write $\img(R) = \{\rho : \pi
\patternrule \rho \in R\}$. We write the reflexive transitive closure of
$\to_R$ as $\tostar_R$. 

We can look at the relation $\to_R$ as a function with its domain being all permutations containing
patterns from the domain of $R$ and its image all permutations containing patterns from the image of
$R$.

An element $x$ in $\Sym$ is said to be in \emph{normal form} under the relation $\to_R$
if there is no $y \in \Sym$ so that $x \to_R y$. A relation $\to_R$ is said to be
\emph{terminating} if there exist no infinite chains $x_0 \to_R x_1 \to_R \cdots$. If
$\to_R$ is terminating then every element has at least one normal form. 

We say a relation is \emph{confluent} if and only if
$$
    y_1 \fromstar x \tostar y_2 \Longrightarrow 
    \exists z. \  y_1 \tostar z \fromstar y_2.
$$
Intuitively, if $\to_R$ is confluent, it implies that every path we can
take using $\to_R$, from a given permutation, will end up in the same place. 

If a system is both terminating and confluent, then every element has a one and
only one normal form.

\begin{corollary}
    If $R$ is a terminating and confluent pattern-rewriting system, then the set of
    $\dom(R)$-avoiding permutations in $S_n$, which we also write $S_n(\dom(R))$, is a complete set of
    representatives for $S_n / \equiv_R$, the set of equivalence classes of $\equiv_R$.
\end{corollary}

It is easy to create confluent systems from terminating one but it is rather difficult to find
terminating systems. Later in this paper we give a lemma that makes it easy to determine if a given
system is terminating. We then show that it works for the majority of equivalence relations defined
by partitions of $\Sym_3$.

We can generate rewrite systems from families of equivalences. Each pair $\{\alpha, \beta \}$ in the
canonical form can either be made into the rule $\alpha \patternrule \beta$ or $\beta \patternrule
\alpha$. The number of different rewrite systems we can generate from $J$ is $2^{|J|}$.

\begin{definition}
  We say a pattern-rewriting system $R$ is \emph{disjoint} if $\dom(R) \cap \img(R) = \emptyset$.
\end{definition}

\begin{example}
  The pattern equivalence $\{ \alpha, \beta \}$ induces an equivalence relation.
  That equivalence relation is equal to the equivalence closure of two different pattern-rewriting
  relations: The relation induced by $\{ \alpha \patternrule \beta \}$ or the relation induced by
  $\{ \beta \patternrule \alpha \}$. Both systems are disjoint.

  The pattern equivalence $\{ \alpha, \beta, \gamma \}$ also induces an equivalence relation.
  That equivalence relation induced by this set of interchangable patterns is equal to the
  equivalence closure of six different pattern-rewriting
  relations: The relation induced by $\{ \alpha \patternrule \beta, \beta \patternrule \gamma \}$,
  $\{ \alpha \patternrule \beta, \alpha \patternrule \gamma \}$, etc. We see that the first example is
  not disjoint while the second one is. If we only look at disjoint systems then we get
  $\{ \alpha \patternrule \gamma, \beta \patternrule \gamma \}$,
  $\{ \alpha \patternrule \beta, \gamma \patternrule \beta \}$,
  $\{ \beta \patternrule \alpha, \gamma \patternrule \alpha \}$.
  They all share the property of having a single pattern in their image.
\end{example}

\subsection{Increasing statistics}
It is rather difficult to determine if a system is terminating. Claesson introduced a method to
do so involving \emph{increasing statistics}.
A \emph{statistic} is any function $f : \Sym \to \mathbb{N}$. We call $f$ an increasing
statistic with respect to $R$ if $a \to_R b$ implies that $f(a) < f(b)$.

\begin{lemma}
    Let $R$ be a pattern-rewriting system. If there exists an increasing
    statistic with respect to $R$, then $\to_R$ is terminating.    
\end{lemma}

\begin{definition}
    Let the statistic $\Sigma_\tau(a)$ be the sum of the positions of occurrences of
    the pattern $\tau$ in the permutation $a$. We call this statistic the sigma statistic
    parameterized by the pattern $\tau$.
    \cite{claesson:2021}.
\end{definition}

Let's take one example from Claesson's paper.
\begin{example}
    Let's look at the equivalence $\{ \{ 12, 21 \} \}$. We can select the system $\{
        21 \to 12 \}$. This relation induced by the system is terminating since the statistic
    $\Sigma_{12}$ is increasing on the relation. It's not
    confluent though, since $321 \to 231$ and $321 \to 312$. So we add the rule
    $231 \patternrule 312$ to the system. The relation is still terminating and by Lemma
    3.3 in Claesson's paper we also have a confluent system. We then end up with
    the system 
    \[
        \{ 21 \patternrule 12, 231 \patternrule 312 \}
    \]
    which is the result we are focusing on in this article.

    By his machinery we find that 
    \[
        \sum_{n \geq 1} \Sym_n(21, 231) = \sum_{m \geq 0} m!(x-x^2)^m
    \]
    which is the joint distribution of the equivalence classes defined by 
    $\{ \{ 12, 21 \} \}$, as shown by Stanley \cite{stanley:2012}.
\end{example}

We will look more at increasing statistics and show that if systems fulfil a certain property called
\emph{overlap invariance} then they are terminating due to a certain increasing statistic.

\section{Finding terminating and confluent systems}
Claesson only found systems corresponding to equivalence relations induced by a
single set of equivalent patterns. We introduce a new type of statistic which can determine the
termination of rewriting systems generated from families with multiple equivalences.
\subsection{Overlap invariance}
Let us begin by generalizing the sigma statistic Claesson introduced in his paper.
\begin{definition}
    A sigma statistic parameterized by a set of patterns $\Pi$ is defined as
    $$
        \Sigma_\Pi = \sum_{\pi \in \Pi} \Sigma_{\pi}.
    $$
\end{definition}
alongside reiterating the definition of $\Ocal$:
\begin{definition}
    For $\sigma, \tau \in T$ define the set of permutations $\Ocal(\sigma,
    \tau)$ by stipulating that $\pi \in \Ocal(\sigma, \tau)$ if and only if
    \begin{enumerate}
    \item $\sigma$ is a proper prefix of $\pi$,
    \item $\tau$ is a proper suffix of $\pi$ and
    \item $|\pi| < |\sigma| + |\tau|$.
    \end{enumerate}
\end{definition}

Then, we will define a cluster \emph{overlap} and the function $\ovl$.
\begin{definition}
  When two patterns or permutations cluster we call the bit they share their \emph{overlap}.
  For permutations $a$ and $b$, we define $\ovl(a,b)$ as the maximum length of their overlaps if they 
  cluster. If they do not cluster then we set $\ovl(a,b)=0$, the length of their overlap is zero.

  For example $\ovl(213, 123) = 1$ since $\Ocal(213, 123) = \{ 21\underline{3}45 \}$. For $\ovl(123, 123) = 2$
  because $\Ocal(123, 123) = \left\{ 1\underline{23}4, 12\underline{3}45 \right\}$. Finally
  $\ovl(123, 321)=0$ since $123$ and $321$ do not cluster.
\end{definition}

We can then define \emph{overlap invariance}.
\begin{definition}
    Let $R$ be a pattern-rewriting system with rules of length $n$. If
    \[
        \pi_j = \rho_j, n-m \leq j \leq n \  \textnormal{where}
        \ m = \max\nolimits_{\tau \in \img(R)} \ovl(\pi, \tau)
    \]
    for each $\pi \patternrule \rho \in R$ then we say the system is \emph{overlap invariant}.
\end{definition}

The term overlap invariance refers to the fact that if a domain of a rule in the system clusters
with some pattern in the image of the system, then that rule does not alter the overlap. That
specific type of pattern overlap is invariant.

\begin{example}
  The system
  \begin{align*}
    (1) 123 \patternrule 213 \\
    (2) 312 \patternrule 132
  \end{align*}
  is overlap invariant. The left hand side of (1) clusters with the right hand side of (2) and 
  $123_3 = 213_3$. The left hand side of (2) clusters with neither 213 nor 132.

  We see that we call a system overlap invariant since if we apply a rule on a domain pattern that clusters
  with an image pattern to its right, then we do not alter the overlap of the two patterns.
  Therefore the image pattern to the right stays the same and clusters with a new image pattern.
  
  Let us look at the following permutation, which has the occurrences of $\img(R)$ marked bold and
  occurrences of $\dom(R)$ underlined.
  \[
    \underline{12\bm{3}}\bm{54}
  \]
  We apply the rule $123 \patternrule 213$ to get the permutation
  \[
    \bm{21\bm{3}}\bm{54}
  \]
  Since the system is overlap invariant the occurrence of the pattern 132 was left unchanged while
  an occurrence of 213 was added.
\end{example}

This leads us to a new lemma:

\begin{lemma}
  If $R$ is overlap invariant and disjoint then $\Sigma_{\img(R)}$ is an increasing statistic with respect to $R$. 
\end{lemma}
\begin{proof}
  Since the system is disjoint there is no rule $a \patternrule b$ in $R$ with $a, b \in \img(R)$.
  That rule would not move an image pattern to the right, just replace it with another one, which is
  not enough to increase the statistic value of the resulting permutation.

    Let $a \to_R b$. Then there is a rule $\pi \patternrule \rho$ in
    $R$ so that $u \pi v \to_R u \rho v$, where $a = u \pi v$ and $b =
    u \rho v$. 

    If elements of $\img(R)$ is fully contained in either $u$ or $v$ then we
    can safely ignore them. If an element appears in $u\pi$, at the end of $u$ and the beginning of
    $\pi$ then $u \rho$ will have a higher sigma value since the position of a pattern in $\img(R)$ will
    effectively move to the right. If an element appears in $\pi v$,
    then by overlap invariance it will also appear in $\rho v$.
    So
    \[
        \Sigma_{\img(R)}(a) < \Sigma_{\img(R)}(b)
    \]

\end{proof}

And since an increasing statistic under a system $R$ implies termination of the rewrite relation
induced by $R$ then we have as a corollary:

\begin{corollary}
    Let $R$ be overlap invariant and disjoint. Then $R$ induces a terminating pattern-rewriting relation.
\end{corollary}

\subsection{Making a terminating system confluent}
Lemma $3.3$ in Claesson's paper gives us a convenient test to tell whether a system is
confluent. 

What we need to show is a system is locally confluent, that is
\begin{lemma}[Lemma 3.3 \cite{claesson:2021}]
    Let $R$ be a terminating pattern rewriting system. Let 
    \begin{equation*}
        \Ocal(R) = \cup_{\pi_1, \pi_2} \Ocal(\pi_1, \pi_2),
    \end{equation*}
    where the union is over all pairs $(\pi_1, \pi_2) \in \dom(R) \times \dom(R)$. If, for all
    $\pi \in \Ocal(R)$,
    \begin{equation*}
        \rho_1 \from \pi \to \rho_2 \Longrightarrow \exists \sigma. \rho_1 \fromstar \sigma \tostar
        \rho_2,
    \end{equation*}
    then $\to_R$ is confluent.
        
\end{lemma}

Using this test and a method for adding rules we devise an algorithm. The method we are using is the
naive one, we simply find shortest cluster of the domain of the terminating system that has more than two
normal forms under the system and we add a rule that joins these two normal forms together, making
sure that application of the rule would give a higher sigma statistic.

Below is pseudocode for the algorithm.

\lstset{basicstyle=\footnotesize\ttfamily, frame=single, breaklines=true,
mathescape=true}
\begin{minipage}{\linewidth}
\begin{lstlisting}[title={Algorithm 1}]
confluentize(sys):
    in: a terminating system, sys
        increasing statistic that proves termination of sys, f
    out: a terminating and confluent system or failure

    if sys is confluent then
        return sys

    s := choose shortest cluster in O(R) with more than one normal form
    rules := { a -> b | a, b are normal forms of s, f(a -> b) > 0 }

    if rules = $\emptyset$ then
        return failure

    rule := rule from rules chosen by the maximum f(rule)

    return algo(sys + rule)
\end{lstlisting}
\end{minipage}

Let us provide a proof that these added rules which make the system confluent do
not make the system non-terminating.
 
We have to make sure that the rule must add images or moves them to the right and does not ruin
occurrences of any of the images present in the source permutation.

\[
\begin{tikzcd}  
    & \bm{img}\mathbf{domdom}\bm{img} \arrow[dl] \arrow[dr] & \\
    img\underline{\mathbf{img}\mathrm{dom}}\bm{img} \arrow[bend
    right=30, mapsto]{rr} & & \bm{img}\underline{\mathrm{dom}\mathbf{img}}\bm{img}
\end{tikzcd}
\]

Here img represents a pattern in the image of a system, and dom a pattern in the domain of a system.
If either is bold then that means that it is an actual occurrence of the pattern but if it is not
then the occurrence has been ruined by an application of a rule from the system. The italicized
img's at the beginning and the end of the pattern represent possible occurrences of img around the
pattern, just to show that they are not ruined either when applying this new rule.

%%% fixit
By casual induction we see that even if we take domain from the confluence rules
then they will also be overlap invariant.

We see that the system is still terminating under the original $\img(R)$ as
that a confluence rule is also overlap invariant under $\img(R)$.

\subsection{Examples}
To illustrate the utility of the methods we have introduced we will look at two
examples

\begin{example}
Let $J=\{\{ 132, 312 \}, \{ 213, 231, 321 \}\}$. This family can generate 6
different disjoint rewrite systems. We test each one for overlap invariance and one that
matches is 
\[
    \begin{matrix}
        123 \patternrule 312 \\
        132 \patternrule 312 \\
        231 \patternrule 321 \\
    \end{matrix}
\]
We then run our confluentzation algorithm on this terminating system
and get the system
\[
    \begin{matrix}
        123 \patternrule 312 \\
        132 \patternrule 312 \\
        231 \patternrule 321 \\
        3124 \patternrule 1423 \\
        31254 \patternrule 12534
    \end{matrix}
\]
This system is both confluent and terminating, which is our end goal.
With that kind of system $R$ we can go on enumerating
permutations that avoid $\dom(R)$ with Claesson's method and get the joint distribution
\[
  \sum_{m \geq 0} m! (-3x^3 + x)^m.
\]
 
\end{example}

\begin{example}
Let $J = \{ \{ 123, 132, 312 \}, \{ 231, 321 \} \}$ for which there is a overlap invariant, hence
terminating, and confluent pattern-rewriting system
\[
    \begin{matrix}
        123 \patternrule 312 \\
        132 \patternrule 312 \\
        231 \patternrule 321 \\
        3124 \patternrule 1423 \\
        31254 \patternrule 12534
    \end{matrix}
\]
Then the joint distribution is
\[
  \sum_{m \geq 0} m! (-3x^3 + x)^m.
\]

\end{example}

\section{Overview of results}
We can apply the above mentioned results by testing all disjoint rewrite systems of a
family of equivalences for overlap invariance. If it is overlap invariant then it is terminating and
we can make it both terminating and confluent system by adding some rules.

We can look at a set partition as a set of disjoint sets, therefore a set partition of a set of
patterns is a set of disjoint sets of patterns, which is the same as a family of
equivalences.

For example, let us look at one set partition of $\Sym_3$, namely 
\[
  \left\{ \left\{ 123, 132 \right\}, \left\{ 213 \right\}, \left\{ 231, 312 \right\}, \left\{ 321
  \right\} \right\}.
\]
We can ignore the singleton sets, they simply represent that its member is equivalent to itself. We
then get the family of equivalences
\[
  \left\{ \left\{ 123, 132 \right\}, \left\{ 231, 312 \right\} \right\}.
\]


To test the efficacy of the overlap invariance idea we look at every set partition of 
\[
  \Sym_3 = \left\{ 123, 132, 213, 231, 312, 321 \right\}.
\]
Those are 203, of those are 155 families of equivalences from which we can derive overlap
invariant systems.
Calculating the joint distribution of those systems gives us nine different Wilf classes.

\begin{definition}
    We can give a family of equivalences $J$ a size function which is the same
    as the cardinality of a canonical form of the set of equivalences.

    By this definition 
    \begin{align*}
      & | \{ \{ 123, 321 \}, \{ 312, 231, 213 \} \}| =  \\
      & | \{ \{ 123, 321 \}, \{ 312, 231 \}, \{ 312, 213 \} \} | =  3
    \end{align*}
\end{definition}

Let us collate all the results from Appendix B into a table to show how the
different generating functions are distributed.

The joint distribution is always on the form $\sum_{m \geq 0}m!(x+C_J(x))^m$ where $C$
is the cluster generating function corresponding to the system $J$.

\begin{center}
\begin{tabular}{c|c|c|c}
    $|J|$ & $C_J(x)$ & count & OEIS seq. \\
    \hline
    0 & $0$ & 1 & A000142 \\
    1 & $-x^3$ & 13 & A212580 \\
    2 & $-2x^3$ & 46 & A212432 \\
    2 & $\frac{-x^3-x^4}{x^2+x+1}$ & 2 & A212581 \\
    3 & $-3x^3$ & 45 & \emph{Not present} \\
    3 & $\frac{-2x^3-2x^4-x^5}{x^2+x+1}$ & 10 & A212433 \\
    4 & $-4x^3$ & 4 & \emph{Not present} \\
    4 & $\frac{-4x^3-3x^4-2x^5}{x^2+x+1}$ & 4 & \emph{Not present}\\
    4 & $\frac{-4x^3-2x^4}{x^2+x+1}$ & 1 & \emph{Not present} \\ 
    \hline
    \hline
    5 & $\frac{-5x^3-3x^4-x^5}{x^2+x+1}$ & 1 & \emph{Not present} \\ 
\end{tabular}
\end{center}

If we parameterize each $C_J$ with it's corresponding $|J|$ value then we have
three different types of cluster generating functions.

Those types are $C'_J(x) = -|J|x^3$, $C''_J(x) = \frac{-|J|x^3-(|J|-1)x^4-(|J|-2)x^5}{x^2+x+1}$
and $C'''_J(x) = \frac{-|J|x^3-(|J|-2)x^4-(|J|-4)x^5}{x^2+x+1}$

For $C'''$ we only generated one system whereas Claesson had generated a system for
$J = \{\{ 123, 132, 213, 231, 312, 321 \}\}$ which cluster generating function
is also $C'''_J$. We include it in our table to show that a pattern holds.

Thus, these partitions of $\Sym_3$ are split into three different types
of Wilf-classes, which we can describe by $C'$, $C''$ and $C'''$ and the size
of the partition.

\subsection{Programming}
There were three small programs written for the project: mp for generating the
systems from the equivalence relations, mp for taking the domain of these
systems and calculating the generating functions and finding if their sequences
were found in the OEIS, and finally tester which calculates the first 6 values
of the sequences manually. mp and tester were written in Haskell and mp in Sage Math. 

\section{Problems and questions}

There are a few questions that are left unanswered by this paper.
Are there any other Wilf classes that appear for relations defined by partitions on $\Sym_3$? There are
still approxmimately 50 relations for which we have not found a corresponding Wilf class. 

Not every relation has a corresponding overlap invariant system, but still have terminating and
confluent systems. An example of those would be the relation induced by $\{ \{ 213, 132
\} \}$. Its rewrite systems ${\{ 213 \patternrule 132 \}}$ and ${\{
132 \patternrule 213 \}}$ are not overlap invariant. 
But for $\{ 213 \patternrule 132 \}$ sigma parameterized by $21$ is an increasing statistic. We
should be able to find these sigma statistics parameterized by smaller patterns algorithmically.
That should be possible by looking at all Hertzsprung patterns that appear in a left- and righthand
side of a rule. Then you can tally if a patterns either appears, disappears or moves when applying the rule.

There are also some questions that arise regarding the relation between a family of equivalences $J$ and
the cluster generating function $C_J$.
Would similar formulas to $C', C''$ and $C'''$ appear when we would
generate systems from partitions of $\Sym_4$? 
Or more generally, 
can we derive a way to create a cluster generating function directly from a
given family of equivalences? 

On a seperate note,
is there an algorithm that makes all systems confluent?

Let us end with one conjecture and one open problem:
\begin{conjecture}
    If a partition $J$ on $\Sym_n$ has a system that has no non-self-clusters in
    its domain then its cluster generating function is $x^n-|J|x$.
\end{conjecture}

\begin{openproblem}
    Find a terminating statistic for the rest of the equivalence relations
    defined by partitions of $\Sym_3$.
\end{openproblem}

\section*{Acknowledgments}

\bibliographystyle{plain}
\bibliography{mp}

\clearpage

\begin{appendices}

\clearpage

\newgeometry{left=2cm,right=2cm,top=2cm,bottom=2cm}
\section{Resulting systems}
We categorize the equivalence families/equivalence relations by what their joint
distribution is.
\begin{multicols}{2}
\include{appendix}
\end{multicols}
\end{appendices}

\end{document}
